### 2012-09-22 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
adding load balance example

\section{Preliminaries}
\label{sec:nonlinear_preliminaries}
We formulate the \textit{nonlinear problem} as follows
\citep{knoll_jacobian-free_2004}:
\begin{equation}
  \ve{F}(\ve{u}) = \ve{0}\:,
  \label{eq:nonlinear_problem}
\end{equation}
where $\ve{u}$ is the solution vector and $\ve{F}(\ve{u})$ is the
function of nonlinear residuals. We write the nonlinear system in this
form so that when an exact solution for $\ve{u}$ is achieved, all
residuals evaluate to zero. \textit{Newton's method} is a root finding
algorithm and therefore we can use it to solve
Eq~(\ref{eq:nonlinear_problem}) if we interpret the exact solution
$\ve{u}$ to be the roots of $\ve{F}(\ve{u})$. Newton's method is also
an iterative scheme, and we can generate this scheme by generating the
Taylor expansion of the $k+1$ iterate of the nonlinear residuals about
the $k$ iterate:
\begin{equation}
  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
  \:.
  \label{eq:newton_derivation_1}
\end{equation}
If we ignore the nonlinear terms in the expansion and assert that the
$k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following:
\begin{equation}
  -\ve{F}(\ve{u}^{k}) =
  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})\:.
  \label{eq:newton_derivation_2}
\end{equation}
We note two things in Eq~(\ref{eq:newton_derivation_2}). The first is
that $\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
$\ve{J}(\ve{u})$, of the set of nonlinear residuals and is defined
element-wise as:
\begin{equation}
  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}\:.
  \label{eq:jacobian_def}
\end{equation}
Second, we note that $(\ve{u}^{k+1}-\ve{u}^{k})$ is simply the
\textit{Newton correction} at the $k$ iterate, $\delta \ve{u}^k$. We
can then rerarrange Eq~(\ref{eq:newton_derivation_2}) to define the
Newton iteration scheme for nonlinear problems:
\begin{subequations}
  \begin{gather}
    \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})\\
    \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k\:.
  \end{gather}
  \label{eq:newton_iteration}
\end{subequations}
We then have two distinct steps to perform: the solution of a linear
system to compute the Newton correction where the Jacobian matrix of
the nonlinear equation set is the linear operator, and the application
of the correction to the previous iterate's solution to arrive at the
next iterate's solution. In the asymptotic lmiit, the iterations of
Newton's method will converge the nonlinear residual quadratically
\citep{kelley_iterative_1995}. Convergence criteria is set for
stopping the iteration sequence based on the nonlinear
residual. Commonly the following criteria is used:
\begin{equation}
  ||\ve{F}(\ve{u}^{k})|| < \epsilon ||\ve{F}(\ve{u}^{0})||\:,
  \label{eq:newton_stopping_criteria}
\end{equation}
where $\epsilon$ is a user defined tolerance parameter. Newton's
method is \textit{consistent} in that all components of the nonlinear
functions that describe the physics we are modeling are updated
simulataneously in the iteration sequence with respect to one
another. This is in comparison to \textit{inconsistent} strategies,
such as a pressure correction strategy for solving the Navier-Stokes
equations \citep{pletcher_computational_1997}, where the components of
$\ve{u}$ are updated in a staggered fashion depending on the
particular equations that they are associated with.

\section{Inexact Newton Methods}
\label{sec:newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied.

\subsection{Newton-Krylov Methods}
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that use a Krylov subspace method as the
linear solver for generating the Newton correction. As we investigated
in Chapter~\ref{chap:linear_problem}, Krylov methods are robust and
enjoy efficient parallel implementations. Furthermore, their lack of
explicit dependence on the operator make them easier to implement than
other methods.

In most cases, the Jacobian is difficult to form from the difference
equations. For simple nonlinear cases such as the Navier Stokes
equations, the derivatives can be computed and coded, but due to the
complexity of those derivatives and the resulting difference equations
this task can be tedious, error prone, and must be repeated for every
equation set. By recognizing that Krylov methods only need the action
of the operator on the vector instead of the operator itself, the
Jacobian can instead be approximated through various numerical methods
including a difference-based Jacobian-free formulation. Jacobian-Free
methods, and in particular \textit{Jacobian-Free Newton-Krylov} (JFNK)
methods \citep{knoll_jacobian-free_2004}, rely on forming the action
of the Jacobian on a vector as required by the Krylov solver through a
forward difference scheme. In this case, the action of the Jacobian on
some vector $\ve{v}$ is given as:
\begin{equation}
  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
    \ve{F}(\ve{u})}{\epsilon}\:,
  \label{eq:jacobian_free_product}
\end{equation}
where $\epsilon$ is a small number typically on the order of machine
precision. Kelley \citep{kelley_iterative_1995} points out a potential
downfall of this formulation in that if the discretization error in
$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
$\epsilon$, then the finite difference error from
Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.

>>>>

\section{Preliminaries}
\label{sec:nonlinear_preliminaries}
We formulate the \textit{nonlinear problem} as follows
\citep{knoll_jacobian-free_2004}:
\begin{equation}
  \ve{F}(\ve{u}) = \ve{0}\:,
  \label{eq:nonlinear_problem}
\end{equation}
where $\ve{u} \in \mathbb{C}^n$ is the solution vector and
$\ve{F}:\mathbb{C}^N \rightarrow \mathbb{C}^N$ is the function of
nonlinear residuals. We write the nonlinear system in this form so
that when an exact solution for $\ve{u}$ is achieved, all residuals
evaluate to zero. \textit{Newton's method} is a root finding algorithm
and therefore we can use it to solve Eq~(\ref{eq:nonlinear_problem})
if we interpret the exact solution $\ve{u}$ to be the roots of
$\ve{F}(\ve{u})$. Newton's method is also an iterative scheme, and we
can generate this scheme by generating the Taylor expansion of the
$k+1$ iterate of the nonlinear residuals about the $k$ iterate:
\begin{equation}
  \ve{F}(\ve{u}^{k+1}) = \ve{F}(\ve{u}^{k}) +
  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k}) +
  \frac{\ve{F}''(\ve{u}^{k})}{2}(\ve{u}^{k+1}-\ve{u}^{k})^2 + \cdots
  \:.
  \label{eq:newton_derivation_1}
\end{equation}
If we ignore the nonlinear terms in the expansion and assert that the
$k+1$ iterate $\ve{u}^{k+1}$ is that exact solution such that
$\ve{F}(\ve{u}^{k+1}) = \ve{0}$, then we are left with the following:
\begin{equation}
  -\ve{F}(\ve{u}^{k}) =
  \ve{F}'(\ve{u}^{k})(\ve{u}^{k+1}-\ve{u}^{k})\:.
  \label{eq:newton_derivation_2}
\end{equation}
We note two things in Eq~(\ref{eq:newton_derivation_2}). The first is
that $\ve{F}'(\ve{u}^{k})$ is in fact the \textit{Jacobian},
$\ve{J}(\ve{u})$, of the set of nonlinear residuals and is defined
element-wise as:
\begin{equation}
  J_{ij} = \frac{\partial F_i(\ve{u})}{\partial u_j}\:.
  \label{eq:jacobian_def}
\end{equation}
Second, we note that $(\ve{u}^{k+1}-\ve{u}^{k})$ is simply the
\textit{Newton correction} at the $k$ iterate, $\delta \ve{u}^k$. We
can then rerarrange Eq~(\ref{eq:newton_derivation_2}) to define the
Newton iteration scheme for nonlinear problems:
\begin{subequations}
  \begin{gather}
    \ve{J}(\ve{u}) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})\\
    \ve{u}^{k+1} = \ve{u}^k + \delta \ve{u}^k\:.
  \end{gather}
  \label{eq:newton_iteration}
\end{subequations}
We then have two distinct steps to perform: the solution of a linear
system to compute the Newton correction where the Jacobian matrix of
the nonlinear equation set is the linear operator, and the application
of the correction to the previous iterate's solution to arrive at the
next iterate's solution. In the asymptotic lmiit, the iterations of
Newton's method will converge the nonlinear residual quadratically
\citep{kelley_iterative_1995}. Convergence criteria is set for
stopping the iteration sequence based on the nonlinear
residual. Commonly the following criteria is used:
\begin{equation}
  ||\ve{F}(\ve{u}^{k})|| < \epsilon ||\ve{F}(\ve{u}^{0})||\:,
  \label{eq:newton_stopping_criteria}
\end{equation}
where $\epsilon$ is a user defined tolerance parameter. Newton's
method is \textit{consistent} in that all components of the nonlinear
functions that describe the physics we are modeling are updated
simulataneously in the iteration sequence with respect to one
another. This is in comparison to \textit{inconsistent} strategies,
such as a pressure correction strategy for solving the Navier-Stokes
equations \citep{pletcher_computational_1997}, where the components of
$\ve{u}$ are updated in a staggered fashion depending on the
particular equations that they are associated with.

\section{Inexact Newton Methods}
\label{sec:newton_methods}
Inexact Newton methods arise when the Jacobian operator is not exactly
inverted, resulting in an inexact Newton correction as initially
described by Dembo and others \citep{dembo_inexact_1982}. For common
sparse nonlinear systems, which in turn yield a sparse Jacobian
matrix, this situation occurs when conventional iterative methods are
applied. In their definition, Dembo formulated inexact methods such
that they are independent of the linear method used to solve for the
Newton correction and therefore are amenable to use with any linear
solver.

\subsection{Newton-Krylov Methods}
\label{subsec:newton_krylov_methods}
A form of inexact Newton methods, \textit{Newton-Krylov methods} are
nonlinear iterative methods that use a Krylov subspace method as the
linear solver for generating the Newton correction. As we investigated
in Chapter~\ref{ch:linear_problem}, Krylov methods are robust and
enjoy efficient parallel implementations. Furthermore, their lack of
explicit dependence on the operator make them easier to implement than
other methods.

\subsubsection{Choosing a Krylov Method}
\label{subsubsec:krylov_choice}
From their numerical studies, McHugh and Knoll found
that the robustness enjoyed by the long recurrence relations used by
GMRES for the orthogonalization procedure proved to be a better choice
for use with Newton's method
\citep{mchugh_inexact_1993,knoll_newton-krylov_1995}.

\subsubsection{Jacobian-Free Approximation}
\label{subsubsec:jacobian_free_approximation}
In most cases, the Jacobian is difficult to form from the difference
equations. For simple nonlinear cases such as the Navier Stokes
equations, the derivatives can be computed and coded, but due to the
complexity of those derivatives and the resulting difference equations
this task can be tedious, error prone, and must be repeated for every
equation set. By recognizing that Krylov methods only need the action
of the operator on the vector instead of the operator itself, the
Jacobian can instead be approximated through various numerical methods
including a difference-based Jacobian-free formulation. Jacobian-Free
methods, and in particular \textit{Jacobian-Free Newton-Krylov} (JFNK)
methods \citep{knoll_jacobian-free_2004}, rely on forming the action
of the Jacobian on a vector as required by the Krylov solver through a
forward difference scheme. In this case, the action of the Jacobian on
some vector $\ve{v}$ is given as:
\begin{equation}
  \ve{J}(\ve{u})\ve{v} = \frac{\ve{F}(\ve{u} + \epsilon \ve{v}) -
    \ve{F}(\ve{u})}{\epsilon}\:,
  \label{eq:jacobian_free_product}
\end{equation}
where $\epsilon$ is a small number typically on the order of machine
precision. Kelley \citep{kelley_iterative_1995} points out a potential
downfall of this formulation in that if the discretization error in
$\ve{F}(\ve{u})$ is on the order of the perturbation parameter
$\epsilon$, then the finite difference error from
Eq~(\ref{eq:jacobian_free_product}) pollutes the solution.
